---
title: "LA-ICP-MS fission track dating"
author: "Tobias Stephan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LA-ICP-MS fission track dating}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Basics

LA-ICP-MS FT single-grain age calculation using the 'zeta' approach:

$t = \frac{1}{\lambda}~\ln \left(1+\frac{1}{2} \lambda \zeta_{icp} \frac{N_s}{A_s [ ^{238}U/^xX]} \right)$

single grain uncertainties:

$\frac{s[t]}{t} \approx \sqrt{ \left( \frac{s[\zeta_{icp}]}{\zeta_{icp}} \right) ^2 + \left( \frac{s[^{238}U/^xX]}{[^{238}U/^xX]} \right) ^2 + \frac{1}{N_s} }$


The pooled age is

$\overline{t} = \frac{1}{\lambda}~\ln \left( 1 + \frac{1}{2} \lambda \zeta_{icp} \frac{\sum N_s}{\sum A_s [^{238}U/^{x}X]} \right)$


Session 'zeta' is given by

$\zeta_{icp} = \frac{\left( e^{\lambda t} -1 \right) U   }{ \frac{1}{2} \lambda \rho_s}$

and its uncertainties are

$\frac{s[\zeta_{icp}]}{\zeta_{icp}} = \sqrt{ \frac{1}{N_s} + \left( \frac{s[t]}{t}\right)^2 + \left(\frac{s[^{238}U/^{x}X]}{[^{238}U/^{x}X]}\right)^2}$


## Data import

``U`` [U content] is given as the ratio between 238U/43Ca based on cps (for apatite) or  U/Si (zircon). Therefor, we have to rename the columns.

```{r import, warning=FALSE,message=FALSE}
library(readxl)
library(dplyr)

library(laftr)

# load the U concentrations
sample <- read_xlsx("../inst/extdata/example.xlsx", sheet = "sample") %>%
  rename(U = UCa, sU = sUCa)

# load the calibration measurements for zeta factor
zeta.measurements <- read_xlsx("../inst/extdata/example.xlsx", sheet = "standard") %>%
  rename(U = UCa, sU = sUCa)
```

## Zeta factor of the ICP-session

`zeta_ICP()` by giving the calibration measurements and the used standard:
```{r zeta1}
session.zeta <- zeta_ICP(zeta.measurements, spotsize = 40, mineral = "apatite", standard = "Dur")
```
It returns a list with the original data:

```{r zeta2}
head(session.zeta$data) %>% as_tibble()
```

...  the area, track densities, and the individual zetas:
```{r zeta3}
head(session.zeta$results) %>% as_tibble()
```

... and the final zeta factor (with uncertainty)
```{r zeta4}
session.zeta$zeta %>% as_tibble()
```


## Age calculation

`age_ICP()` using the just calculated zeta factor for calibration

```{r res}
result <- age_ICP(sample, spotsize = 40, zeta = session.zeta$zeta)
```

returns a list containing the original dataset:
```{r res.data}
head(result$data) %>% as_tibble()
```

.. the individual ages:
```{r ages}
head(result$ages) %>% as_tibble()
```

... the pooled age of the sample:
```{r ages.pooled}
result$age.pooled
```

... and the chi-squared statistics:
```{r chisq}
result$stats
```


## Compatibility with IsoplotR

```{r isoplotr}
# library(IsoplotR)
data.iso <- as.fissiontracks(sample, spotsize = 40, zeta = session.zeta$zeta)
ages.iso <- as.fissiontracks(result$ages, ages = TRUE)
```


### KDE

using the functions from the `IsoplotR` package (P. Vermeesch):

```{r kde}
IsoplotR::kde(ages.iso[, 1])
```

or the `laftr` function `geom_aKDE()` using adaptive KDE with `ggplot2`:

```{r ggkde, warning=FALSE, message=FALSE}
library(ggplot2)

data.frame(t = ages.iso[, 1], st = ages.iso[, 2]) %>%
  ggplot(aes(x = t, weight = t / st)) +
  geom_aKDE(aes(y = after_stat(scaled)), fill = "steelblue", alpha = .75) +
  geom_histogram(aes(y = after_stat(ncount)), color = "grey", fill = "grey", alpha = .5) + # weighted histogram
  geom_rug(alpha = 0.5)
```

### Radial plot

```{r radial}
IsoplotR::radialplot(ages.iso)
```

### Central Age

```{r central}
IsoplotR::central(ages.iso)
```

### Weighted mean

```{r w.mean}
IsoplotR::weightedmean(ages.iso, ranked = TRUE)
```

### Peak fitting

```{r peaks}
IsoplotR::peakfit(ages.iso)
```
